---
title: "Alternative Distinctiveness definition"
author: "Rekyt"
date: "04/08/2018"
header-includes:
  - \usepackage{mathtools}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The current definition of functional distinctiveness standardizes compared to all the species considered in the species pool. However, especially when investigating on trait competition and hierarchy, it can be important to scale this distinctiveness relatively to the species present in the community.

## Presence-Absence

Thus an alternative definition of functional distinctiveness can be given, scaled relatively to a given trait range considered (what matters for species are the species within this trait range, not beyond). Our new distinctiveness measure is $D_i(T)$, with $T$ the trait range to consider.

We can imagine three species in a given community positioned on a trait axis:
```{r}
species_df = data.frame(
  species =     c("a", "b", "c"),
  trait_value = c(-1, 0, 0.5)
)

species_distance = dist(c(a = -1, b = 0, c =0.5))

plot(species_df)
species_distance
```
Species $b$ is closer to species $c$ than to species $a$. The greatest distance is observed between species $a$ and species $c$.

$T$ can value from 0 to the maximum observed functional distance among considered species.

### Theoretical definition

The proposed formula for $D_i(T)$ is:

$$
D_{i} (T) = 
  \begin{dcases}
    1 & if~~T < \text{min}(d_{ij}) \quad \text{(trait range smaller than distance to closest neighbor)}\\
    \frac{\sum\limits_{
          \substack{
                j = 1 \\
                j \neq i \\
                d_{ij} \leq T}}^S d_{ij}}{
          \sum\limits_{\substack{
                j = 1 \\
                j \neq i \\
                d_{ij} \leq T}}^S 1}
                 & if~~T \geq \text{min}(d_{ij}) \quad (\text{there is one or more species close to focal species within range of T})
  \end{dcases} 
$$
This would given the following formula for the distinctiveness of species $b$, in the community with species $a$ and $c$:

$$
D_b(T) =
  \begin{dcases}
    1                         & if ~~ T < d_{bc} = 0.5 \\
    d_{bc}                    & if ~~ d_{bc} = 0.5 \leq T < d_{ba} = 1 \\
    \frac{d_{bc} + d_{ba}}{2} & if ~~ T \geq d_{ba} = 1 
  \end{dcases}
$$

### Implementation

Implementing this version of distinctiveness would give:

```{r new_distinctiveness}
alternative_distinctiveness = function(pres_mat, distance_obj, given_T) {
  dist_mat = as.matrix(distance_obj)
  
  kept_sp = funrar:::species_in_common(pres_mat, dist_mat)
  dist_mat = dist_mat[kept_sp, kept_sp, drop = FALSE]
  
  # Correspondance matrix (tracking which species we want to keep)
  corr_dist = dist_mat
  corr_dist[dist_mat > given_T] = 0
  corr_dist[dist_mat <= given_T] = 1
  diag(corr_dist) = 0
  
  di_mat = diag(dist_mat %*% corr_dist) / rowSums(corr_dist)
  di_mat[rowSums(corr_dist) == 0] = 1
  
  dplyr::mutate(tibble::enframe(di_mat, "species", "Di"), T_value = given_T)
}
```
Here we truncate the distance matrix in order to "nullify" the effect of species that are not in specified range.

```{r}
presence_matrix = matrix(rep(1, 3), nrow = 1, ncol = 3,
                         dimnames = list(site = "s1",
                                         species = c("a", "b", "c")))

# Test over a ragne of possible T values
all_T = lapply(seq(0.5, 1.5, length.out = 50),
       function(given_number) alternative_distinctiveness(presence_matrix,
                                                          species_distance,
                                                          given_number))

all_T = dplyr::bind_rows(all_T)
```

We can visualize those relationships:
```{r}
library(ggplot2)

ggplot(all_T, aes(T_value, Di, color = species)) +
  geom_line(size = 2) +
  xlim(0, 1.5) +
  ylim(0, 1.5) +
  labs(x = "Fixed T range",
       y = "Functional Distinctiveness",
       color = "Species")
```


### Using real data

We'll be using the dataset `aravo` from the `ade4` package.

```{r aravo_new_distinct}
library("funrar")

data("aravo", package = "ade4")
# Site-species matrix
mat = as.matrix(aravo$spe)

# Convert matrix to presence-absence matrix
mat[mat > 0] = 1

# Example of trait table
tra = aravo$traits[, c("Height", "SLA", "N_mass")]
# Distance matrix
dist_mat = compute_dist_matrix(tra, metric = "gower")
dist_mat = (dist_mat - min(dist_mat))/diff(range(dist_mat))

# Compute different values of distinctiveness using various ranges
all_ranges = lapply(seq(0, 1, length.out = 50), function(given_range) {
  alternative_distinctiveness(mat, as.dist(dist_mat), given_range)
})

all_ranges = dplyr::bind_rows(all_ranges)
```

Now we can visualize the influence of the trait range value

```{r}
ggplot(all_ranges, aes(T_value, Di, group = species)) +
  geom_line(alpha = 1/3) +
  labs(x = "Minimum Distance Range Considered\n(Trait Range)",
       y = "Functional Distinctiveness")
```

We see that species have close neighbors at different distances (leap from top left to bottom left of the graph). This shows the closet neighbors for each species.

As proposed we can also distinctiveness relative to trait range and see what it changes:
```{r scaled_range}
all_ranges = dplyr::mutate(all_ranges, scaled_Di = ifelse(Di != 1, Di / T_value, Di))

ggplot(all_ranges, aes(T_value, scaled_Di, group = species)) +
  geom_line(alpha = 1/4) +
  labs(x = "Considered Trait Range\n(Functional Distance)",
       y = "Scaled Functional Distinctiveness\n(over trait range)")
```

This reflects the relative (in terms of trait range considered) distinctiveness. Meaning that close species in a small range brings similar distinctiveness than further apart species in larger ranges.

## Considering Abundance

Now that we have a version with presence and absences we can try to think about a alternative version of distinctiveness using abundances.

We need the index to behave the following way: an rare species in trait range should be more distinct than a common species in that range. A species close to a rare species in a trait range but far from a common species should be more distinct than a species close to a common species and far away from a rare one.

The current definition of functional distinctiveness with relative abundances is:

$$

\widetilde{D_i} = \frac{
  \sum\limits_{
      \substack{j = 1 \\
                j \neq i}}^S d_{ij}
  \times Ab_j}{
  \sum\limits_{
      \substack{j = 1 \\
                j \neq i}}^S
  Ab_j}

$$
with $Ab_j$ the relative abundance of species $j$ in given community.

### Theoretical definition

If we try to consider the constraints above we can write the index as follow

$$
D_i(T) =
  \begin{dcases}
  1 & if \\
  \left( \frac{
    \sum\limits_{
      \substack{j = 1 \\
                j \neq i \\
                d_{ij} \leq T}}^S d_{ij} \times Ab_j
  }{
    \sum\limits_{
      \substack{j = 1 \\
                j \neq i \\
                d_{ij} \leq T}}^S Ab_j
  } \right) \times \left(1 - \frac{
    \sum\limits_{
      \substack{j = 1 \\
                j \neq i \\
                d_{ij} \leq T}}^S Ab_j
  }{
    N
  } \right) & if
  \end{dcases}
$$


