---
title: "Other functions of interest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Other functions of interest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(funrar)
```

In addition to single index functions `funrar` proposes several functions to help you compute functional rarity indices. The aim of this vignette is to provide detailed examples of how to use each of them.


## Compute Distinctiveness from Global Pool - `distinctiveness_global()`

While functional distinctiveness has been envisioned initially at local scale, that's what the `distinctiveness()` function computes, it can be useful to compute it at regional scales.
This functions provides exactly that based on a provided functional dissimilarity matrix.


## Get rarity across combinations of traits - `rarity_dimensions()`

Distinctiveness is generally computed through a dissimilarity matrix that itself comes from aggregating differences across multiple traits.

### Different Distinctiveness - `distinctiveness_dimensions()`

### Different Uniqueness - `uniqueness_dimensions()`


## Compute Relative Abundances - `make_relative()`

Function to transform matrix of absolute abundances to relative abundances.
Takes an abundance matrix and returns the same matrix with relative abundances.
